---
title: "2019 BPO Projections"
author: "GLA demography"
date: '`r strftime(Sys.time(), format = "%B %Y")`'
output: html_document
params: 
  x: !r x
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r title}
full_area_name <- ifelse(x$area_name == x$borough_name,
                     paste("the borough of",x$borough_name),
                     paste(x$area_name, "ward in", x$borough_name))

area_title <- ifelse(x$area_name == x$borough_name,
                     paste(x$area_name, "(borough total)"),
                     paste(x$area_name, "ward"))

space <- " "

```

## `r area_title`

This is summary of the outputs of the GLA 2019-based Borough Preferred Option (BPO) projections for `r full_area_name`. These projections use the latest 2019 mid-year small area population estimates and a housing development trajectory specified by colleagues in `r x$borough_name`.

This output is provided as a QA tool and a quick reference to the BPO scenario projections. Full ward-level projections are available to download from the borough BPO hub on the London Datastore. If you have any queries please contact demography@london.gov.uk.

### Migration assumptions
The BPO projections presented here vary in their assumptions about future migration.\ 

`r space`\  

#### Covid period (2020-2022)
In the initial years of the projection there are two migration variants: the **standard** and the **high out migration** variants. Scenarios 1 and 2 use the standard assumptions while scenario 3 uses the high out migration assumptions. For full details of the migration assumptions used see the accompanying BPO documentation.

Broadly speaking, the high out migration variant assumes a scenario in which the impacts of the pandemic on migration are more significant and the recovery to pre-pandemic behaviours is slower. Both domestic and international flows are adjusted with the overall impact being a higher total net outflow of population from London.\ 

`r space`\  

#### Transition period (2023-2027)
During this period migration rates and flows are adjusted incrementally until the long-term level is reached in 2028.\

`r space`\  
 
#### Long-term period (2028-2050)
There are two longer term migration variants: a **standard** and a **low domestic migration** variant. Scenarios 1 and 3 use the standard variant while scenario 2 uses the low migration variant. Again, full details are available in the accompanying BPO documentation.

In the both variants, international flows return to the 10-year average (2010-2019) in 2028.

In the standard variant domestic migration rates return to the 5-year average (2015-2019) in 2028.

In the low migration variant domestic migration returns to a 10-year average (2010-2019) in 2028. The 10-year average is lower than the 5-year average and so produces lower domestic migration flows. In broad terms this means that population will increase at a greater rate as fewer people out-migrate from London.\ 

`r space`\  

## Total population  {.tabset}

### 2010-2050
```{r total_pop}
x$total_pop_chart
```

### 2019-2030
```{r pop_short_term}
x$short_term_chart
```

## {-}

## School-age population  {.tabset}

### primary age
```{r primary_pop}
x$primary_chart
```

### secondary age

```{r secondary_pop}
x$secondary_chart
```
  
## {-}
  

## Age Structure  {.tabset}

### scenario 1
```{r age_structure_2011}
x$age_structure_charts[[1]]
```

### scenario 2
```{r age_structure_2019}
x$age_structure_charts[[2]]
```

### scenario 3
```{r age_structure_2034}
x$age_structure_charts[[3]]
```

## {-}

```{r age_table}
tbl <- x$age_table %>% rename(` ` = age_group)
knitr::kable(tbl, caption = "Population by selected age groups")
```

## Components of Change {.tabset}

The components of population change are births, deaths and net migration. Data up to and including 2019 are adapted from ONS mid-year population estimates. Data for 2020-2050 are modelled and projected by GLA demography.

### births
  
```{r total_births}
x$births_chart
```

### deaths

```{r total_deaths}
x$deaths_chart
```

### migration

```{r net_migration}
x$migration_chart
```

## {-}

## Assumed Development

BPO projections use the GLA housing-led model to reconcile projected trends in population change with future housing capacity. The development trajectory used is specified by the commissioning borough. The chart shows the development trajectory for `r x$area_name` used in the 2019 BPO projections. 
  
```{r assumed_dev_chart}
x$assumed_dev_chart
```

### Notes
Copyright GLA demography 2021.\
Contact demography@london.gov.uk\
BPO projections are not published by the GLA but may be published by the commissioning authority.\
Thanks to Jonathan West at Waltham Forest.\
  
  