---
params:
  set_title: !r paste0("BPO Projection - ",ward_name)
title: "`r params$set_title`"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(flexdashboard)
```


**Projection: `r paste0(bpo_name, " BPO development scenario")`**


This is document provides an overview of the BPO projections for `r ward_name` The projections were run using a ward-level development trajectory specified by the borough.

Three variant projections have been run using the same development trajectory. These variants differ in the rates of domestic migration employed in the model. The variants are as follows:

* High - Migration rates are based on a recent period of high mobility (2016-2018) for the entire projection period.

* Medium - in the short-term (from 2019 to 2024) the rates used are based on the most recent 5 years' domestic flows. For the remainder of the projection, the average of the last 10 years' rates is used.

* Low - in the short-term (from 2019 to 2024) the rates used are based on the most recent 5 years' domestic flows. For the remainder of the projection, the average rates of the period 2009 to 2012 are used. This represents conditions under an economic downturn/recession.

The document is not inteneded to be comprehensive but provieds an at-a-glance overview of some key population indicators and components of change.

The full dataset has been made available in the borough BPO hub on the London Datastore.

Row
------------------------------------------------

### Total population `r ward_name`
```{r total popn, echo=FALSE}
chart_1[[current_ward]]
```

Row
------------------------------------------------


### Primary-age population (5-10) 

```{r primary, echo=FALSE}
chart_primary[[current_ward]]
```

### Secondary-age population (11-15)

```{r secondary, echo=FALSE}
chart_secondary[[current_ward]]
```

Row {data-height=80}
-------------------------------------------------

Row {data-height=40}
-------------------------------------------------

**Population Change**

Row {data-height=60}
-------------------------------------------------

The charts below show population in each year as a proportion of the 2011 population. The borough trajectory is shown for context. A ward with a trajectory greater than that of he borough is projected to grow at a faster rate than the borough average, while lower trajectories indicate slower growth.


Row {.tabset}
---------------------------------------------

### Low migration sceanrio

```{r popn low, echo=FALSE}
chart_change_low[[current_ward]]
```

### Medium migration sceanrio

```{r popn med, echo=FALSE}
chart_change_med[[current_ward]]
```

### High migration sceanrio

```{r popn high, echo=FALSE}
chart_change_high[[current_ward]]
```

Row {data-height=80}
-------------------------------------------------

Row {data-height=40}
-------------------------------------------------
**Age Structure**

Row {data-height=40}
-------------------------------------------------

The age sructure at three points (2011, 2018 and 2033) is shown for each of the three migration scenarios in the charts below. The table provides snapshot detail for six population sub-groups.

Row {.tabset}
---------------------------------------------

### Low migration sceanrio

```{r age strcuture low, echo=FALSE}
chart_age_strcture_low[[current_ward]]
```

### Medium migration scenario

```{r age strcuture med, echo=FALSE}
chart_age_strcture_med[[current_ward]]
```

### High migration sceanrio

```{r age strcuture high, echo=FALSE}
chart_age_strcture_high[[current_ward]]
```

Row
---------------------------------------------
### Population by age group
```{r age group table, echo = FALSE}
knitr::kable(age_table[[current_ward]])
```


Row {data-height=80}
---------------------------------------------

Row {data-height=40}
-------------------------------------------------

**Components**

Row {data-height=60}
-------------------------------------------------

Births and deaths for the period 2011-2018 are actuals taken from ONS outputs. Net migration for the period 2011-2018 is a residual calculated as the different between the starting and ending population in any given, year minus births and plus deaths. 

Row {.tabset}
---------------------------------------------

### Net migration
```{r net migration, echo=FALSE}
chart_net_migration[[current_ward]]
```

### Births
```{r births, echo=FALSE}
chart_births[[current_ward]]
```

### Deaths
```{r deaths, echo=FALSE}
chart_deaths[[current_ward]]
```

Row {data-height=80}
---------------------------------------------

Row {data-height=40}
-------------------------------------------------

**Assumed Development**

Row {data-height=60}
-------------------------------------------------

Assumed development is taken from the submitted BPO development trajectory. For year 2012-2018 recorded development from the London Development Database (LDD) is used. For other years, where no data has been provied, unit values are takenfrom the 2017 SHLAA.

Row
---------------------------------------------

### Assumed development `r ward_name`
```{r development, echo=FALSE}
chart_dev[[current_ward]]
```

Row
---------------------------------------------

*Past births, deaths and population data are copyright ONS. Projected data are copyright GLA Demography 2020.*
