---
params:
  set_title: !r paste0(ward_data$ward_name, ", ", ward_data$borough_name)
title: "`r params$set_title`"
output:
  html_document:
    number_sections: no
    includes:
      before_body: header.html
      after_body: footer.html
    css: "style.css"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(plotly)
```

```{r functions, echo = FALSE}
nice_number <- function(x){format(round(x,0), big.mark=",")}
space <- '&nbsp;'
ward_name <- ward_data$ward_name
```


# `r paste0(ward_data$dev_trajectory, " development trajectory")`

`r space`


The Borough Preferred Option (BPO) projections are a special set of housing-led population projections produced by the GLA for London local authorities. The BPO projections require an input housing trajectory and typically two sets of BPO projections are produced: one set using a housing development trajectory specified by the borough for this purpose, and one set using a trajectory adapted by the GLA from the 2017 Strategic Housing Land Availability Assessment (SHLAA). The projections shown here were produced using a `r ward_data$dev_trajectory` development trajectory.

## Ward geography

The 2021-based BPO projections (produced in early 2023) are produced on 2022 wards boundaries.

## Population base and backseries

In December 2022 ONS released the 2021 Mid-Year Estimate (MYE) for local authorities in England and Wales. This release constitutes a break in the official MYE series as it is based on rolling forward 2021 census data rather than the 2020 MYE.

In order to have a consistent backseries to work with the GLA have modelled a local authority level series between the accepted data points 2011 and 2021. This LA series has been modelled down to output area and then re-aggregated up to 2022 ward boundaries. The result is a consistent ward-level backseries for the period 2011-2021.

However, this process may mean that ward populations in 2021 differ in size and distribution from census 2021 outputs.

## Adittional data

The GLA have produced a modelled set of local authority level births for the year to mid-2022. These combine publicly-available GP registrations and births data to now-cast births ahead of official statistical releases. The GLA modelled estimates for 2022 are incorporated into the small area modelling.

## Model operation

The BPO projections use the GLA flexible area model to produce ward-level population projections. 

The model operates by calculating a population based solely on trends and then separately one based on converting available dwelling stock into population through the application of average household size (AHS). These two populations are then reconciled in the model to produce a final population which accounts for both current trends and relative changes in housing availability.

## Migration scenarios

The 2021-based BPO projections include two scenario projections for each development trajectory:

* **Scenario 1**: Ward migration rates are calculated by applying the average of the last 5 years' rates to projected population. The population, births and deaths are constrained to the 2021-based 5-year migration trend projection.

* **Scenario 2**: Ward migration rates are calculated by applying the average of the last 10 years' rates to projected population. The population, births and deaths are constrained to the 2021-based 10-year migration trend projection.


### Constraining

The constraining process has the effect of transferring the overarching characteristics of the trend projection, and its detailed migration assumptions, to the ward-level projection.

The trend projections assume the same level of international migration in 2022 (a figure based on ONS data on UK-level international flows for that year). International migration levels then diverge as the variants move towards their long-term levels. These levels are achieved in 2025 and then remain constant to the end of the projection. 

Domestic migration levels in the constraining trend projections are calculated by applying averaged rates to projected population in each year of the projection.

`r space`

```{r chart_layout, echo = FALSE}


chart_layout <- function(p, y_min_value, y_max_value){
  
  
  p %>%          #2 rectangles for different periods
    layout(
      shapes = list(
        list(type = "rect",
             fillcolor = "#44aad5",
             line = list(width = 0),
             opacity = 0.15,
             x0 = 2021, x1 = 2025,
             y0 = y_min_value, y1 = y_max_value,
             layer = "below"),
        
        list(type = "rect",
             fillcolor = "#44aad5",
             line = list(width = 0),
             opacity = 0.05,
             x0 = 2025, x1 = 2042,
             y0 = y_min_value, y1 = y_max_value,
             layer = "below")),
      
      #Annotation at the top of the rectangles
      annotations = list(
        list(x = 2023,
             y = 0.93,
             text = "Short-term\nperiod",
             xref = "x",
             yref = "paper",
             xanchor = 'center',
             yanchor = 'top',
             showarrow = FALSE,
             font=list(size=11)),
        
        list(x = 2032,
             y = 0.93,
             text = "Long-term period",
             xref = "x",
             yref = "paper",
             xanchor = 'center',
             yanchor = 'top',
             showarrow = FALSE,
             font=list(size=11)),
        
        #Source text in the bottom left
        list(x = 1,
             y = -0.10, #position of text adjust as needed 
             text = "Source: GLA 2021-based BPO projections", 
             showarrow = F,
             xref='paper',
             yref='paper', 
             xanchor='right',
             yanchor='auto',
             xshift=0,
             yshift=0,
             font=list(size=11, color="grey"))
      )
    )
}

```

## Population {.tabset}

### Total population

```{r total_pop, echo = FALSE}
population <- ward_data$population 

p <- population %>% 
  data.frame() %>% 
  plot_ly(x=~year, y=~value, color=~scenario,
          colors = c("#6da7de",
                     "#9e0059"),
          type = "scatter",
          mode = "lines+markers", 
          line = list(shape = "spline"),
          text = ~scenario,
          hovertemplate = paste("%{text}: %{y:,.0f}",
                                "<extra></extra>"),
          width = 1000,
          height = 600) %>% 
  layout(title = paste0("Total population - ",ward_name ,"\n2021-based BPO scenario projections"),
         xaxis = list(title = "",
                      showgrid = FALSE),
         yaxis = list(title = "persons",
                      showgrid = TRUE),
         legend = list(orientation = "h"),
         margin = list(b=50),
         hovermode = "x unified")

y_min_value <- min(population$value)
y_max_value <- max(population$value)

chart_layout(p, y_min_value, y_max_value)

```

### Primary age

```{r primary, echo = FALSE}
primary <- ward_data$primary

p <- primary %>% 
  data.frame() %>% 
  plot_ly(x=~year, y=~value, color=~scenario,
          colors = c("#6da7de",
                     "#9e0059"),
          type = "scatter",
          mode = "lines+markers", 
          line = list(shape = "spline"),
          text = ~scenario,
          hovertemplate = paste("%{text}: %{y:,.0f}",
                                "<extra></extra>"),
          width = 1000,
          height = 600) %>% 
  layout(title = paste0("Primary population (4-10) - ", ward_name, "\n2021-based BPO scenario projections"),
         xaxis = list(title = "",
                      showgrid = FALSE),
         yaxis = list(title = "persons",
                      showgrid = TRUE),
         legend = list(orientation = "h"),
         margin = list(b=50),
         hovermode = "x unified")

y_min_value <- min(primary$value)
y_max_value <- max(primary$value)

chart_layout(p, y_min_value, y_max_value) 


```

### Secondary age

```{r secondary, echo = FALSE}
secondary <- ward_data$secondary

p <- secondary %>% 
  data.frame() %>% 
  plot_ly(x=~year, y=~value, color=~scenario,
          colors = c("#6da7de",
                     "#9e0059"),
          type = "scatter",
          mode = "lines+markers", 
          line = list(shape = "spline"),
          text = ~scenario,
          hovertemplate = paste("%{text}: %{y:,.0f}",
                                "<extra></extra>"),
          width = 1000,
          height = 600) %>% 
  layout(title = paste0("Secondary population (11-15) - ", ward_name, "\n2021-based BPO scenario projections"),
         xaxis = list(title = "",
                      showgrid = FALSE),
         yaxis = list(title = "persons",
                      showgrid = TRUE),
         legend = list(orientation = "h"),
         margin = list(b=50),
         hovermode = "x unified")

y_min_value <- min(secondary$value)
y_max_value <- max(secondary$value)

chart_layout(p, y_min_value, y_max_value) 

```

### Age Structure

```{r age_structure, echo = FALSE}
age_structure <- ward_data$age_structure

p <- age_structure %>% 
  data.frame() %>% 
  plot_ly(x=~age, y=~value, color=~year,
          colors = c("#6da7de",
                     "#9e0059",
                     "#dee000",
                     "#000000"),
          type = "scatter",
          mode = "lines+markers", 
          line = list(shape = "spline"),
          text = ~year,
          hovertemplate = paste("%{text}: %{y:,.0f}",
                                "<extra></extra>"),
          width = 1000,
          height = 600) 

y_min_value <- min(age_structure$value)

p %>% 
  layout(title = paste0("Age structure - ", ward_name, "\n2021-based BPO scenario projections"),
         xaxis = list(title = "",
                      showgrid = FALSE),
         yaxis = list(title = "persons",
                      showgrid = TRUE),
         legend = list(orientation = "h"),
         margin = list(b=50),
         hovermode = "x unified",
         
         annotations = list(
           #Source text in the bottom left
           list(x = 1,
                y = -0.10, #position of text adjust as needed 
                text = "Source: GLA 2021-based BPO projections", 
                showarrow = F,
                xref='paper',
                yref='paper', 
                xanchor='right',
                yanchor='auto',
                xshift=0,
                yshift=0,
                font=list(size=11, color="grey"))
         )
         
         
  )
```


`r space`

## Migration {.tabset}

The BPO scenarios differ in their migration assumptions. Scenario 1 applys an average of the last 5 years' migration rates while Scenario 2 applys an average of the last 10 years' rates.

### In Migration

```{r in_migration, echo = FALSE}

inflow_data <- ward_data$in_migration

p <- inflow_data %>% 
  plot_ly(x=~year, y=~value, color=~scenario,
          colors = c("#6da7de",
                     "#9e0059"),
          type = "scatter",
          mode = "lines+markers", 
          line = list(shape = "spline"),
          text = ~scenario,
          hovertemplate = paste("%{text}: %{y:,.0f}",
                                "<extra></extra>"),
          width = 1000,
          height = 600)  %>% 
  layout(title = paste0("In Migration - ", ward_name, "\n2021-based BPO scenario projections"),
         xaxis = list(title = "",
                      showgrid = FALSE),
         yaxis = list(title = "persons",
                      showgrid = TRUE),
         legend = list(orientation = "h"),
         margin = list(b=50),
         hovermode = "x unified")


y_min_value <- min(inflow_data$value)
y_max_value <- max(inflow_data$value)

chart_layout(p, y_min_value, y_max_value) 

```

### Out Migration

```{r out_migration, echo = FALSE}

outflow_data <- ward_data$out_migration

p <- outflow_data %>% 
  plot_ly(x=~year, y=~value, color=~scenario,
          colors = c("#6da7de",
                     "#9e0059"),
          type = "scatter",
          mode = "lines+markers", 
          line = list(shape = "spline"),
          text = ~scenario,
          hovertemplate = paste("%{text}: %{y:,.0f}",
                                "<extra></extra>"),
          width = 1000,
          height = 600)  %>% 
  layout(title = paste0("Out Migration - ", ward_name, "\n2021-based BPO scenario projections"),
         xaxis = list(title = "",
                      showgrid = FALSE),
         yaxis = list(title = "persons",
                      showgrid = TRUE),
         legend = list(orientation = "h"),
         margin = list(b=50),
         hovermode = "x unified")

y_min_value <- min(outflow_data$value)
y_max_value <- max(outflow_data$value)

chart_layout(p, y_min_value, y_max_value)         

```

### Net Migration

```{r net_migration, echo = FALSE}

netflow_data <- ward_data$net_migration

p <- netflow_data %>% 
  plot_ly(x=~year, y=~value, color=~scenario,
          colors = c("#6da7de",
                     "#9e0059"),
          type = "scatter",
          mode = "lines+markers", 
          line = list(shape = "spline"),
          text = ~scenario,
          hovertemplate = paste("%{text}: %{y:,.0f}",
                                "<extra></extra>"),
          width = 1000,
          height = 600) %>% 
  layout(title = paste0("Net Migration - ", ward_name, "\n2021-based BPO scenario projections"),
         xaxis = list(title = "",
                      showgrid = FALSE),
         yaxis = list(title = "persons",
                      showgrid = TRUE),
         legend = list(orientation = "h"),
         margin = list(b=50),
         hovermode = "x unified")

y_min_value <- min(netflow_data$value)
y_max_value <- max(netflow_data$value)

chart_layout(p, y_min_value, y_max_value)  
```

## Fertility
Mid-year estimate births are available up to and including the year to mid-2021. Births for 2022 are caluculated from GP registrations data and years following are calculated by applying projected fertility rates to the population at single-year-of-age.

The 2021-based BPO projections use fertility rates based on trending the most recent years' data forward. In past years the model has used an average of past rates as standard with the trend option provided on request. The trend of declining in births observed in London since the peak in 2012 leads us to believe that a trended fertility rate provides the most reasonable starting point for projecting future births in the 2021-based BPO projections.

`r space`

```{r births, echo = FALSE}

births_data <- ward_data$births

p <- births_data %>% 
  plot_ly(x=~year, y=~value, color=~scenario,
          colors = c("#6da7de",
                     "#9e0059"),
          type = "scatter",
          mode = "lines+markers", 
          line = list(shape = "spline"),
          text = ~scenario,
          hovertemplate = paste("%{text}: %{y:,.0f}",
                                "<extra></extra>"),
          width = 1000,
          height = 600)  %>% 
  layout(title = paste0("Births - ", ward_name, "\n2021-based BPO scenario projections"),
         xaxis = list(title = "",
                      showgrid = FALSE),
         yaxis = list(title = "persons",
                      showgrid = TRUE),
         legend = list(orientation = "h"),
         margin = list(b=50),
         hovermode = "x unified")

y_min_value <- min(births_data$value)
y_max_value <- max(births_data$value)

chart_layout(p, y_min_value, y_max_value)

```

## Mortality

Mid-year estimate deaths are available up to and including the year to mid-2021. Deaths for 2022 and years following are calculated by applying projected fertility rates to the population at single-year-of-age.

Deaths from Covid-19 are assumed to be accounted for by the normal application of mortality rates.

`r space`

```{r deaths, echo = FALSE}

deaths_data <- ward_data$deaths

p <- deaths_data %>% 
  plot_ly(x=~year, y=~value, color=~scenario,
          colors = c("#6da7de",
                     "#9e0059"),
          type = "scatter",
          mode = "lines+markers", 
          line = list(shape = "spline"),
          text = ~scenario,
          hovertemplate = paste("%{text}: %{y:,.0f}",
                                "<extra></extra>"),
          width = 1000,
          height = 600) %>% 
  layout(title = paste0("Deaths - ", ward_name, "\n2021-based BPO scenario projections"),
         xaxis = list(title = "",
                      showgrid = FALSE),
         yaxis = list(title = "persons",
                      showgrid = TRUE),
         legend = list(orientation = "h"),
         margin = list(b=50),
         hovermode = "x unified")

y_min_value <- min(deaths_data$value)
y_max_value <- max(deaths_data$value)

chart_layout(p, y_min_value, y_max_value)


```


## Housing development data

#### Past development

For the years 2012-2019 the development assumptions are taken from the London Development Database (LDD). This data is used in place of borough-supplied data except where the borough has explicitly requested that the LDD not be used.

#### Projected development

For the years 2020-2041, the model requires an input development trajectory.

Where boroughs have not provided development data, or where the provided trajectory does not cover the entire period, the projections use data from the 2017 Strategic Housing Land Availability Assessment (SHLAA) as a proxy for future housing delivery.

For the years 2020-2024 the SHLAA data is adjusted to account for the impact of the pandemic on housing delivery. This adjustment is based on research undertaken by Savills Estate Agents. Development that fails to come forward in these first years of the trajectory is redistributed to later years.

`r space`

```{r dev_chart, echo = FALSE}
dev_data <- ward_data$dev 

p <- dev_data %>% 
  mutate(source = ifelse(year <= 2019, "past", "projected")) %>% 
  plot_ly(x=~year, y=~dwellings, color = ~source,
          colors = c("#6da7de",
                     "#9e0059",
                     "#dee000"),
          type = "bar",
          text = ~source,
          hovertemplate = paste("<b>%{x}</b><br>",
                                "%{text}: %{y:,.0f}",
                                "<extra></extra>"),
          width = 1000,
          height = 600) %>%  
  
  #Global layout and title details
  layout(title = paste0("Development Assumptions - ", ward_name, "\n2021-based BPO scenario projections"),
         xaxis = list(title = "",
                      showgrid = FALSE),
         yaxis = list(title = "net additional dwellings",
                      showgrid = TRUE),
         legend = list(orientation = "h"),
         margin = list(b=50))

chart_layout(p, 0, max(dev_data$dwellings))
       
```


## Interpreting scenario projections

The complex dynamics involved in population change mean that trends at the London level do not necessarily translate directly to the borough and ward level. Which of the two scenarios shows the highest and lowest growth in a borough may differ from the order at the regional level and ward trajectories may not be in the same order as borough trajectories.

**Why do different migration scenarios produce similar population trajectories?**

Where a development trajectory forecasts particularity high levels of housing growth in a borough the role of migration in affecting the population is tempered. As a result, boroughs with significant projected development are likely to see less variation between the 2 migration scenarios. However, there may be significant differences in age structure between projections
which, at the total population level, look similar.

**Are the BPO projections published anywhere?**

The GLA do not make BPO projections publicly available. They are only accessible to those in the boroughs who have signed-up to have access to the BPO portal on the London Datastore. However, if you would like to publish your projections we have no objection. 


## Further information
Please contact the GLA demography team if you have any questions relating to the BPO projections.
demography@london.gov.uk

For questions about the SRP service or projections please contact the SRP team.
srpservice@london.gov.uk


`r space`
